---
title: "PA2_NOAA Storm Database and the effects of severe weather"
author: "Jon Huber"
date: "March 4, 2019"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Github repo for the Course: [Reproducible Research](https://github.com/jlalhuber/Reproducible_Research_2)
</br>

## 1: Synopsis
The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## 2: Data Processing

### *2.1: Data Loading*

Download the raw data file and extract the data into a dataframe.Then convert to a data.table
```{r DataLoading, cache = TRUE}
library("data.table")
library("ggplot2")
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
path <- getwd()
download.file(fileUrl, destfile = paste0(path,'/repdata%2Fdata%2FStormData.csv.bz2'))
storm.df <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
# Converting data.frame to data.table
storm.dt <- as.data.table(storm.df)
```

### *2.2: Examining Data*
Look at column names
```{r ColumnNames}
colnames(storm.dt)
```

Look at a summary of the data
```{r data dimensions}
str(storm.dt)
```
### *2.3: Data Subsetting*

Subset the dataset on the parameters of interest by removing the columns we don't need for clarity. 
```{r DataSubsetting, results="hide"}
# Finding columns to remove
cols2Remove <- colnames(storm.dt[, !c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP")])
# Removing columns
storm.dt[, c(cols2Remove) := NULL]
```

Only use data where fatalities or injuries occurred.  
```{r Data_Injuries_Fatalities}
storm.dt <- storm.dt[(EVTYPE != "?" & 
             (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE"
                                                                            , "FATALITIES"
                                                                            , "INJURIES"
                                                                            , "PROPDMG"
                                                                            , "PROPDMGEXP"
                                                                            , "CROPDMG"
                                                                            , "CROPDMGEXP") ]
```

### *2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)*

Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.
```{r CorrectingExponents, results="hide"}
# Change all damage exponents to uppercase.
cols <- c("PROPDMGEXP", "CROPDMGEXP")
storm.dt[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]
# Map property damage alphanumeric exponents to numeric values.
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)
# Map crop damage alphanumeric exponents to numeric values
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)
storm.dt[, PROPDMGEXP := propDmgKey[as.character(storm.dt[,PROPDMGEXP])]]
storm.dt[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]
storm.dt[, CROPDMGEXP := cropDmgKey[as.character(storm.dt[,CROPDMGEXP])] ]
storm.dt[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```

### *2.5: Making Economic Cost Columns*
Create for total Property Cost and Crop Cost for each row of data
```{r EconomicCostColumns}
storm.dt <- storm.dt[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### *2.6: Calcuating Total Property and Crop Cost*
Create summary of Total Property and Crop Cost by EVTYPE
```{r TotalPropertyCropCost}
totalCost.dt <- storm.dt[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
totalCost.dt <- totalCost.dt[order(-Total_Cost), ]
totalCost.dt <- totalCost.dt[1:10, ]
print(totalCost.dt)
```

### *2.7: Calcuating Total Fatalities and Injuries*
Create summary of Total Fatalities and Injuries by EVTYPE
```{r TotalFatalitiesInjuriesCalc}
totalInjuries.dt <- storm.dt[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
totalInjuries.dt <- totalInjuries.dt[order(-FATALITIES), ]
totalInjuries.dt <- totalInjuries.dt[1:10, ]
print(totalInjuries.dt)
```

## 3: Results

### *3.1: Events that are Most Harmful to Population Health*

Melting data.table so that it is easier to put in bar graph format 
```{r HealthResults}
bad_stuff <- melt(totalInjuries.dt, id.vars="EVTYPE", variable.name = "bad_thing")
head(bad_stuff, 5)
```

```{r healthChart}
# Create chart
healthChart <- ggplot(bad_stuff, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
healthChart = healthChart + geom_bar(stat="identity", aes(fill=bad_thing), position="dodge")
# Format y-axis scale and set y-axis label
healthChart = healthChart + ylab("Frequency Count") 
# Set x-axis label
healthChart = healthChart + xlab("Event Type") 
# Rotate x-axis tick labels 
healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
healthChart = healthChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))
healthChart
```

### *3.2: Events that have the Greatest Economic Consequences*

Melting data.table so that it is easier to put in bar graph format 
```{r EconConsequences}
econ_consequences <- melt(totalCost.dt, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_consequences, 5)
```

```{r econChart}
# Create chart
econChart <- ggplot(econ_consequences, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")
# Format y-axis scale and set y-axis label
econChart = econChart + ylab("Cost (dollars)") 
# Set x-axis label
econChart = econChart + xlab("Event Type") 
# Rotate x-axis tick labels 
econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))
econChart
```